Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-06-05 concurrency/CONCUR 2023/response.txt

Thank you for your time and consideration!

A general note: several reviewers mentioned that they were disappointed in the limitations of the template approach, or wanted to see more evidence that it was effective in practice. In fact, we believe that one of our contributions is finding limitations that were not apparent in the original research [10], and will need to be addressed before putting the approach into practice on a larger scale.

Responses to Reviewer 1's questions:

"Moreover, the paper only gives one example... For the second template (the give-up template) it only discusses why this cannot be directly applied to the BST structure."

Actually, we have also verified the give-up template (for lookup and insert) on the BST. The text at the end of section 4.1 was only meant to imply that we had to break the template interface slightly in order to do so, not that we were unable to. So we do have an example where m = 2.

"Can you give a more explicit discussion of the challenges to use the template approach for C programs?"

"How do your correctness proofs compare with those for the flow interface templates?"

"Can you make your questions about the boundaries of the template/data structure division more explicit? And do you have ideas on how to address them?"

"Are there other examples which can be verified in your approach, e.g. other data structures for lock coupling, or data structures to which multiple templates can be applied?"

"How specific is your change to C? Could it also be used to reason about e.g. Java programs?"

Responses to Reviewer 3's questions:

"I was unable to understand how the updates of the abstract state of the CSS are obtained from the node-level updates at the linearization points. For instance, how do you obtain Node(m[k |-> v]) from Line 18 in Figure 8?"

"In the earlier work, the keysets are tied to the graph structure using an inductive invariant... What is the corresponding mechanism used in this paper?"

"I think the criticism of the template approach based on the proof of the delete operation is perhaps a bit overblown..."

(maybe we just agree with this one?)

Responses to Reviewer 4's questions:

"What exactly is the challenge in realizing this using VST as opposed to other systems like Iris?"

"While you say you'd like to verify structures such as Masstree and Wormhole, it's not clear whether you think the template-based approach would work, especially given your reservations for even "delete" for a bst."

### 2023-08-23 concurrency/CPP 2024/paper.tex
initial change for CPP24

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle
\usepackage{txfonts}
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{parcolumns}
\usepackage{iris}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{Node}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref}}}
\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
\newcommand{\infp}{\ensuremath{\mathsf{Node\_ref}}}

@@

\title{Compositional Verification of Concurrent C Programs with Search Structure Templates} %keep working on this

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://dnguye96.people.uic.edu} }{dnguye96@uic.edu}{https://orcid.org/0000-0002-6810-897X}{%(Optional) author-specific funding acknowledgements
}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Lennart Beringer}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~eberinge} }{eberinge@cs.princeton.edu}{https://orcid.org/0000-0002-1570-3492}{}

\author{William Mansky}{University of Illinois at Chicago, USA\and \url{https://mansky.lab.uic.edu/} }{mansky1@uic.edu}{https://orcid.org/0000-0002-5351-895X}{}

\author{Shengyi Wang}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~shengyiw} }{shengyiw@cs.princeton.edu}{https://orcid.org/0000-0002-2286-8703}{}

\authorrunning{Duc Than Nguyen, Lennart Beringer, William Mansky, and Shengyi Wang} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Duc Than Nguyen, Lennart Beringer, William Mansky, and Shengyi Wang} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011074.10011099.10011692</concept_id>
<concept_desc>Software and its engineering~Formal software verification</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Formal software verification}

\keywords{concurrent separation logic, fine-grained locking, logical atomicity, theorem proving,
	Verified Software Toolchain, Iris} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

\funding{This work was funded in part by NSF Grant \#1811894.}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

%don't say ``real'' so much
\begin{abstract}
%a bit more of a lead-in?
Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort. In this paper, we implement the template approach in the Verified Software Toolchain (VST), and use it to prove correctness of C implementations of fine-grained concurrent data structures. This involves translating code, specifications, and proofs to the idiom of C and VST, and gives us another look at the requirements and limitations of the template approach.
We encounter several questions about the boundaries between template and data structure, as well as some common data structure operations that cannot naturally be decomposed into templates. Nonetheless, the approach appears promising for modular verification of real-world concurrent data structures.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Krishna et al.~proposed concurrent search structure templates~\cite{templates} as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree). The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. %\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

\subsection*{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeted at an implementation language (C) rather than a toy language or algorithm-level representation. %do we want to say this part?

@@

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as \\$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$. Its concurrent counterpart could be specified as 
$$\forall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle$$
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

The core of a search structure template is a \lstinline{traverse} function that navigates a data structure using the chosen synchronization mechanism. A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well. The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters. We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}
There are two things worth noting about this data structure. First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently. Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as 
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *left, *right;} node;
\end{lstlisting}
while a linked list might use
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
typedef struct node {int key; void *value; struct node_t *next;} node;
\end{lstlisting}
This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component. Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.

@@

The first template we consider is lock coupling (also called hand-over-hand locking), in which threads use the locks on each node to prevent interference from other threads during traversal. Each thread always holds at least one lock, and acquires the lock on the next node before releasing its current lock, ensuring that other threads cannot invalidate the ongoing search.
Figure \ref{traverse_lock} shows the lock-coupling \lstinline{traverse} function as presented by Krishna et al.~(Figure \ref{traverse_lock_a}) and our corresponding C implementation (Figure \ref{traverse_lock_b}). The C implementation uses a struct
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct pn {struct node_t *p; struct node_t *n;} pn;
\end{lstlisting}
to mimic the pair of nodes \lstinline{(p, n)} returned by the functional implementation, where \lstinline{n} is the current node and \lstinline{p} is its parent. The template relies on one function provided by the underlying data structure, namely \lstinline{findNext}, which is used to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$. In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a Boolean and, if a next node is found, modifies \lstinline{pn->n}.

@@

\begin{figure}[h]
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
			\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
			\label{traverse_lock_a}	
		\end{subfigure}
		\\ \\ 
		\renewcommand{\thesubfigure}{c}% New fixed/manual numbering
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
			\caption{The \lstinline{insert} method of the lock-coupling template algorithm written in C}
			\label{insert_lock}	
		\end{subfigure}
	\end{subfigure}\qquad
	\renewcommand{\thesubfigure}{b}% New fixed/manual numbering
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms 
		%\wm{I like the layout here, but it would make more sense if \lstinline{traverse} was b and \lstinline{insert} was c}
	}
	\label{traverse_lock}
\end{figure}

@@

We prove correctness of the template by showing that \lstinline{traverse}, \lstinline{insert}, etc. meet logically atomic specifications describing their effects on the data structure. These specifications are defined in terms of assertions $\nodeboxrep$, representing a client thread's handle to the data structure, and $\treerep$, representing the data structure's abstract state. Holding $\nodeboxrep(n)$ for a node $n$ should allow us to acquire the lock on $n$, which in turn gives us access to the contents of the node. Formally, we need to know that the following triples hold:
\begin{align*}
&{\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m \right\rangle}\ \texttt{acquire}(n\texttt{->lock})\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\\
&{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \texttt{release}(n\texttt{->lock})\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
\end{align*}
In other words, $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node. 
%In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).

>>>>

%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%\usepackage{txfonts}
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{parcolumns}
\usepackage{iris}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{Node}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Node\_ref }}}
\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
\newcommand{\infp}{\ensuremath{\mathsf{Node\_ref }}}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Compositional Verification of Concurrent C Programs with Search Structure Templates}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Duc Than Nguyen}
%\authornote{Both authors contributed equally to this research.}
\email{dnguye96@uic.edu}
\orcid{}
\affiliation{%
  \institution{University of Illinois Chicago}
  \streetaddress{}
  \city{Chicago}
  \state{Illinois}
  \country{USA}
  \postcode{}
}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Concurrent search structure templates are a technique for separating the verification of a concurrent data structure into concurrency-control and data-structure components, which can then be modularly combined with no additional proof effort. In this paper, we implement the template approach in the Verified Software Toolchain (VST), and use it to prove correctness of C implementations of fine-grained concurrent data structures. This involves translating code, specifications, and proofs to the idiom of C and VST, and gives us another look at the requirements and limitations of the template approach.
We encounter several questions about the boundaries between template and data structure, as well as some common data structure operations that cannot naturally be decomposed into templates. Nonetheless, the approach appears promising for modular verification of real-world concurrent data structures.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
\label{sec:introduction}
Krishna et al. ~\cite{templates} proposed concurrent search structure templates as a method for proving the correctness of concurrent data structures \emph{compositionally}, separating the proof of a concurrent access method (e.g., optimistic concurrency, hand-over-hand locking, forwarding via internal links) from the proof of the underlying data structure (e.g., linked list, hashtable, B-tree). The concurrency ``templates'' are verified parametrically over data structure operations, and the data structure operations are verified without any reference to concurrency. In theory, this could allow us to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns, and immediately obtain $n \times m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns work only for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. %\wm{Maybe say more about the idea that they only built a proof of concept, and there are still important details not worked out/only tested on ad-hoc examples.}

@@

\subsection*{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent} ~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeted at an implementation language (C) rather than a toy language or algorithm-level representation. %do we want to say this part?

@@

\begin{figure*}

\end{figure*}

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$. Its concurrent counterpart could be specified as 
\begin{mathpar}
	\forall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \\ \texttt{push}(v)\  \\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

The core of a search structure template is a \lstinline{traverse} function that navigates a data structure using the chosen synchronization mechanism. A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well. The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters. We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
	typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}
There are two things worth noting about this data structure. First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently. Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as 
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *left, *right;} 
	node;
\end{lstlisting}
while a linked list might use
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *next;
	} node;
\end{lstlisting}
This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component. Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.

@@

The first template we consider is lock coupling (also called hand-over-hand locking), in which threads use the locks on each node to prevent interference from other threads during traversal. Each thread always holds at least one lock, and acquires the lock on the next node before releasing its current lock, ensuring that other threads cannot invalidate the ongoing search.
Figure \ref{traverse_lock} shows the lock-coupling \lstinline{traverse} function as presented by Krishna et al.~(Figure \ref{traverse_lock_a}) and our corresponding C implementation (Figure \ref{traverse_lock_b}). The C implementation uses a struct
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
	typedef struct pn {struct node_t *p; struct node_t *n;} pn;
\end{lstlisting}
to mimic the pair of nodes \lstinline{(p, n)} returned by the functional implementation, where \lstinline{n} is the current node and \lstinline{p} is its parent. The template relies on one function provided by the underlying data structure, namely \lstinline{findNext}, which is used to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$. In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a Boolean and, if a next node is found, modifies \lstinline{pn->n}.

@@

\begin{figure*}[h]
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
			\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
			\label{traverse_lock_a}	
		\end{subfigure}
		\\ \\ 
		\renewcommand{\thesubfigure}{c}% New fixed/manual numbering
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
			\caption{The \lstinline{insert} method of the lock-coupling template algorithm written in C}
			\label{insert_lock}	
		\end{subfigure}
	\end{subfigure}\qquad
	\renewcommand{\thesubfigure}{b}% New fixed/manual numbering
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms 
		%\wm{I like the layout here, but it would make more sense if \lstinline{traverse} was b and \lstinline{insert} was c}
	}
	\label{traverse_lock}
\end{figure*}

@@

We prove correctness of the template by showing that \lstinline{traverse}, \lstinline{insert}, etc. meet logically atomic specifications describing their effects on the data structure. These specifications are defined in terms of assertions $\nodeboxrep$, representing a client thread's handle to the data structure, and $\treerep$, representing the data structure's abstract state. Holding $\nodeboxrep(n)$ for a node $n$ should allow us to acquire the lock on $n$, which in turn gives us access to the contents of the node. Formally, we need to know that the following triples hold:

\begin{mathpar}
	{\color{blue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{acquire}\left(n\texttt{->lock}\right)\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m\right\rangle}\\
	{\color{blue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ m \right\rangle}\ \\ \texttt{release}(n\texttt{->lock})\ \\ {\color{blue}\left\langle \nodeboxrep(n) \ |\ \treerep\ m\right\rangle}
\end{mathpar}

In other words, $\nodeboxrep(n)$ is sufficient to guarantee that node $n$ is in the abstract state of the data structure and its lock protects associated resources $\mathsf{R}(n)$, the \emph{lock invariant} for the node. 
%In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{n}).

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
		\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
		\right\rangle
	}\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  n', v.
		\begin{array}{l} \texttt{pn} \mapsto \left(n', n'\right) \ \ast \\ \nodeboxrep\left(n'\right) \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
			\\ 
			(\mathsf{if} \ \mathit{res} \ \\ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
			\\ \mathsf{else} \ \mathsf{node\_contents}\left(n', (\texttt{k}, v), \mathsf{range}\right))
		\end{array}
		\ \Bigg | \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.
